---
title: 'Applied Statistics: HW1'
author: "Mikhail Ushakov"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(vegan)
library(psych)
library(reshape2)
library(ggrepel)
library(ggsci)
```


# 1 EDA

## Load data

```{r}
data("BCI")
data("BCI.env")
```

## Explore species counts

NAs

```{r}
sum(is.na(BCI))
```
The data consists of 50 rows representing 1 hectare plot and 225 columns with tree species counts on this plot.

Summary statistics of species counts for each plot

```{r}
describe(t(BCI))
```

Make long df for convenience


```{r}
df <- cbind(plot = rownames(BCI), BCI)
df_long <- melt(df, id.vars = c("plot"), variable.name = "species", value.name = "counts")
```

Plot tree counts distribution

```{r}
rowSums(BCI) %>%  hist(xlab = "Plot trees counts", main = "Plot trees counts distribution")
```

Top 5 most abundant species by median

```{r}
df_long %>% 
  group_by(species) %>% 
  summarise(Mean = mean(counts), Median = median(counts)) %>% 
  arrange(desc(Median)) %>% 
  top_n(5)
```


```{r}
df_long %>% 
  group_by(plot) %>% 
  filter(counts > 0) %>% 
  count() %>% 
  pull(n) %>% 
  hist(
    main='Number of unique species distribution',
    xlab='Number of unique species',
    ylab='Number of plots'
    )
```

## Explore plot environmental variables

NAs

```{r}
sum(is.na(BCI.env))
```

```{r}
head(BCI.env)
```

The second dataset contains 0 variables describing each plot from the first dataset

```{r}
str(BCI.env)
```

There are two variables with information about location of plots.
Also there is an information about elevation and precipitation, but it is equal for all plots.
Age.cat contains info about forest age category, but despite it has 2 factors, there is only one plot with "c3" level.
"Geology" factory has only 1 level.

Thus, the only useful varaibles are Habitat, Stream and EnvHet.


```{r}
BCI.env %>% as.data.frame() %>% group_by(Habitat) %>% count()
```
```{r}
BCI.env %>% as.data.frame() %>% group_by(Stream) %>% count()
```

# Ordination

## Calculate Bray-Curtis distances between plots

```{r}
dist <- vegdist(BCI, method = "bray")
```

## Scree plot

```{r message=FALSE, warning=FALSE, echo=FALSE}
NMDS.scree <- function(x) { #where x is the name of the data frame variable
  plot(rep(1, 10), replicate(10, metaMDS(x, autotransform = F, k = 1, trace = FALSE)$stress), xlim = c(1, 10),ylim = c(0, 0.30), xlab = "# of Dimensions", ylab = "Stress", main = "NMDS stress plot")
  for (i in 1:10) {
    points(rep(i + 1,10),replicate(10, metaMDS(x, autotransform = F, k = i + 1, trace = FALSE)$stress))
  }
}
NMDS.scree(dist)
```


There is no good reflection point, so let's take the number of dimensions of 2

## NMDS

```{r message=FALSE, warning=FALSE, echo=FALSE}
set.seed(7)

NMDS2 <- metaMDS(BCI, k = 2, trymax=100, trace = F, autotransform = F, distance="bray")

```

## Simple ordination plot

```{r}
ordiplot(NMDS2, display = "sites", type="t", cex=0.7)
```

```{r}
ef <- envfit(NMDS2, BCI, permu = 999)
plot(NMDS2, type="t", display="sites")
plot(ef, p.max = 0.001, cex=0.5)
```

## Ordination plot with variables

```{r}
plot(NMDS2, type="n")
points(NMDS2, display="sites", pch = c(1, 2, 3, 4, 5) [as.numeric(BCI.env$Habitat)],col = c("black", "red") [as.numeric(BCI.env$Stream)])
legend(
  "topright",
  legend = c(levels(BCI.env$Habitat)),
  pch = c(1, 2, 3 ,4 ,5),
  title = "Habitat"
)

legend(
  "topleft",
  legend = c(levels(BCI.env$Stream)),
  pch = 1,
  title = "Stream",
  col = c("black", "red")
)

legend(
  "bottomleft",
  paste0("stress = ",round(NMDS2$stress, 3), cex = 1)
)
```

Hard to say that plots are well separated by "Habitat", but it denitiely discriminates some groups.
Seems that OldLow forests can have diversity similar to swamps and OldSlope forests. The second explanation is that it is the most abundant type in the dataset.


```{r}
mds_env <- data.frame(NMDS2$points)
mds_env$habitat <- as.factor(BCI.env$Habitat)
mds_env$envhet <- BCI.env$EnvHet
mds_env$envhet_size <- mds_env$envhet
mds_env$plot <- rownames(mds_env)
```

```{r}
mds_env %>% 
  ggplot(aes(MDS1, MDS2, label = plot, col = habitat)) +
  geom_text_repel(show.legend = F) +
  geom_point(aes(size = envhet)) + 
  scale_color_startrek() +
  theme_classic()
```

The same but more accurate picture we get with ggplot, showing now also heterogenity.
We see that OldHigh and OldLow forests have big difference in heterogeneity and don't see it in OldSlope forests. Young forests and Swamps have big heterogeneity.