---
title: 'Applied Statistics: HW3'
author: "Mikhail Ushakov"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(readxl)
library(vegan)
library(ggplot2)
library(dplyr)
```


# Load and preprocess data

```{r message=FALSE, warning=FALSE}
df <- read_excel('Sleepy lizard.xlsx')
df <- df %>% select(
  Tot_WBC,
  Het_ABS,
  Lym_ABS,
  'H:L Ratio',
  Mon_ABS,
  OthG_ABS,
  LBSI,
  Treatment,
  Habitat,
  Connectivity
)
df$Treatment <-  as.factor(df$Treatment)
df$Habitat <- as.factor(df$Habitat)
df$Connectivity <-  as.factor(df$Connectivity)
```

Data frame with only numeric variables


```{r message=FALSE, warning=FALSE}
df_num <- df %>% select(-c(Treatment, Connectivity, Habitat))
```

# Normalization

```{r message=FALSE, warning=FALSE}
boxplot(df_num)
```

Let's perform centering with log-normalization

```{r message=FALSE, warning=FALSE}

df_norm <- log(df_num + 1)
center <- function(x){
  x - mean(x, na.rm = TRUE)
}

df_cent <- t(apply(df_norm, 1, center))

df_norm <- apply(t(df_cent), 1, center)
```

```{r message=FALSE, warning=FALSE}
boxplot(df_norm)
```

# PCA

```{r}
pca <- prcomp(df_norm)
pca_df <- df %>% select(Treatment, Habitat)
pca_df <- cbind(pca_df, pca$x[, 1:2])
```

```{r message=FALSE, warning=FALSE}
pca_df %>% 
  ggplot(aes(PC1, PC2, col=Habitat, shape=Treatment)) +
    geom_point()
```

# Analysis

## Check dispersion

Let's check equal dispersion in groups before using PERMANOVA

```{r message=FALSE, warning=FALSE}
dist <- vegdist(df_norm, method  = "euclidean")
PCO <- betadisper(dist, df$Treatment)
plot(PCO)
anova(PCO)
```

```{r}
PCO <- betadisper(dist, df$Connectivity)
plot(PCO)
anova(PCO)
```

```{r}
PCO <- betadisper(dist, df$Habitat)
plot(PCO)
anova(PCO)
```

Seems that there is no significant difference in those groups, although Habitat groups look not very nice



## Task 1 

### Blood composition and different landscapes

### Different landscapes

```{r}
adonis2(df_norm ~df$Treatment, method='euclidian')
```

### Habitats from highly modified landscapes

```{r}
df_hab <- df_norm[df$Treatment == 2, ]
hab <- df %>% filter(Treatment == 2)
adonis2(df_hab ~ hab$Habitat, method='euclidian')
```

### Degree of connectivity

```{r}
adonis2(df_hab ~ hab$Connectivity, method = "canberra")

```
## Task 2

```{r}
adonis2(df_hab ~ hab$Habitat + hab$Connectivity, method = "euclidean")
```

# Conclusion

Task 1 results show that:

a. There is a significant difference in blood composition between lizards from different landscapes.
b. Blood composition doesn't differ between lizards from different habitats in highly modified landscapes. 
c. Degree of connectivity doesn't affect the blood composition of lizards from highly modified landscape.

Task 2 results show that there is no evidence to suggest that there is a difference between different combinations of Habitats and Connectivity variables levels.

